---
title: "Principal Component Analysis Report: Fulani_and_Reference_HD_DB"
date: "`r Sys.Date()`"
output: html_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
---

## PCA Fulani including reference populations

We have run SMARTPCA on all Fulani samples and selected reference
populations covering genetic variability within Sahel region as well as
north Africa, Europe and Middle East. The PLINK database used is
`Fulani_and_Reference_HD_DB`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "~/h3a-fulani/")

library(knitr)
library(kableExtra)
library(tidyverse)
library(umap)
library(cluster)
library(ggrepel)
library(ggbeeswarm)
library(wesanderson)
library(RColorBrewer)
library(patchwork)

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

### Distribution of Fulani and reference populations on the map and distribution within regions

```{r color_scales}

c_pal <- c("#FDBB84","#FDBB84","#FDBB84",
"#FC8D59","#FC8D59","#FC8D59",
"#EF6548","#EF6548","#EF6548",
"#D7301F","#D7301F","#D7301F",
"#B30000","#B30000","#B30000",
"#7F0000","#7F0000","#7F0000")

col_palette1 <- c(
  rep(brewer.pal(n = 9, name = 'OrRd')[c(4:9)],4)[c(1:18)],
  rep(brewer.pal(n = 9, name = 'YlGn')[c(4:9)],4)[c(1:19)],
  rep(brewer.pal(n = 11, name = 'BrBG')[c(1:4)],4)[c(1:14)],
  rep(brewer.pal(n = 9, name = 'Greys')[c(4:9)],4)[c(1:15)])

col_palette2 <- wes_palette("Cavalcanti1", 66, type = "continuous")

col_palette3 <- c(wes_palette("GrandBudapest1", 18, type = "continuous"),
                 wes_palette("Cavalcanti1", 19, type = "continuous"),
                 wes_palette("Royal1", 14, type = "continuous"),
                 wes_palette("Darjeeling2", 15, type = "continuous"))
```

```{r fulani_map}

#populations$population_label <- reorder(populations$population_label, populations$fulani)

populations$population_label <- factor(populations$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

world <- ne_countries(scale = "medium", returnclass = "sf")

col_palette <- col_palette1

plot_map <- ggplot(data = world) +
    geom_sf(color = "black", fill="white") +
    xlab("Longitude") + ylab("Latitude") +
    theme_bw() +
    coord_sf(xlim = c(-20.0, 50.0), ylim = c(0.0, 55), expand = TRUE) +
    geom_point(data = populations, 
               aes(x=longitude, y=latitude, color=population_label, shape=population_label), 
               size=2.5, stroke = 0.6) +
    scale_colour_manual(values = col_palette,name="Population") +
    scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
    theme(legend.text = element_text(size=7),
        legend.title = element_text(size=10,face="bold"),
        legend.spacing.x = unit(0.2, 'cm'),
        legend.spacing.y = unit(0.2, 'cm'),
        legend.key.height=unit(0.3,"cm"),
        legend.key.size = unit(0.2,"cm"),
        plot.title = element_text(hjust = 0, vjust=2)) +
    guides(color = guide_legend(ncol=2)) +
    ggtitle("Location of all used Fulani and reference populations")

plot_map
```

```{r}

pdf("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/map.pdf",width=10,height=10)
plot_map
dev.off()
```

Summary of populations by region.

```{r}

population_names <- unique(pca$population_label)

read_tsv("data/samples/comparative_data_all_populations.tsv") %>%
  mutate(fulani=case_when(grepl("Fula", population_label) ~ 1,
                          grepl("Halpularen", population_label, ignore.case = TRUE) ~ 1,
                          region=="West Africa" ~ 2,
                          region=="Central Africa" ~ 3,
                          region=="East Africa" ~ 4,
                          region %in% c("Central Europe","North Europe","South Europe") ~ 7,
                          region=="Middle East" ~ 6,
                          region=="North Africa" ~ 5)) %>%
  filter(population_label %in% population_names) %>%
  group_by(fulani) %>%
  count()
```

## PCA without downsampling of Fulani set

```{r load data, include=TRUE}

data="data/pca-results/Fulani_and_Reference_HD_DB-170351-160222/Fulani_and_Reference_HD_DB.pca.evec"

read_table(data, 
           col_names = FALSE, n_max=1)[2:11] %>% unlist(., use.names=FALSE) -> eigen

eigen <- tibble(pc = c(1:10), eigen=eigen)

p <- ggplot(eigen, aes(x=pc, y=eigen)) +
      geom_line(col="grey") +
      geom_point(col="dark blue") +
      theme_bw() +
      ylab("Eigen Value") +
      xlab("Principal Component")

p
```

#### Processing of the PCA data

```{r load_data, include=FALSE}

pc <- as.character(paste0("PC",c(1:10)))

read_table(data, 
           col_names = FALSE, skip=1) %>%
  dplyr::select(c(1:11)) %>%
  dplyr::rename_at(vars(as.character(paste0("X",c(1:11)))),
            ~c("id",pc)) %>% 
  tidyr::separate(id, into=c("population_label","sample_id"), sep=":") -> pca


## Read populations data
population_names <- unique(pca$population_label)

#color_palette <- brewer.pal(n = 8, name = "Dark2")
#color_palette <- wes_palette("Darjeeling1", 5, type = "discrete")

read_tsv("data/samples/comparative_data_all_populations.tsv") %>%
  filter(population_label %in% population_names) %>%
  mutate(fulani=case_when(grepl("Fula", population_label) ~ 1,
                          grepl("Halpularen", population_label, ignore.case = TRUE) ~1,
                          region=="West Africa" ~ 2,
                          region=="Central Africa" ~ 3,
                          region=="East Africa" ~ 4,
                          region %in% c("Central Europe","North Europe","South Europe") ~ 7,
                          region=="Middle East" ~ 6,
                          region=="North Africa" ~ 5)) -> populations

pca %>% 
  dplyr::group_by(population_label) %>%
  dplyr::summarize(n=length(sample_id),
            PC1=mean(PC1),
            PC2=mean(PC2),
            PC3=mean(PC3),
            PC4=mean(PC4),) %>%
  dplyr::inner_join(populations %>% 
    dplyr::select(population_label, longitude, latitude, lifestyle2, region, country, fulani), 
             by=c("population_label" = "population_label")) -> pca_by_pop

pca %>% 
  dplyr::inner_join(populations %>% 
    dplyr::select(population_label, longitude, latitude, lifestyle2, region, country, fulani), 
             by=c("population_label" = "population_label")) -> pca_with_pop


```

### PC1 x PC2

Given strong sampling of Fulani populations with respect to the
reference panel, Fulani-specific genetic variability was captured by
PC2.

```{r pca_1_2}

#pca_with_pop$population_label <- reorder(pca_with_pop$population_label, pca_with_pop$fulani)

pca_with_pop$population_label <- factor(pca_with_pop$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca12 <- ggplot(pca_with_pop, aes(x=PC1, y=PC2, color=population_label, shape=population_label)) +
  geom_point() +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  theme(legend.position = "none",
        legend.spacing.x = unit(0.2, 'cm'),
        legend.spacing.y = unit(0.3, 'cm'),
        legend.key.size = unit(0.3,"cm"),) +
  guides(color = guide_legend(ncol=4))

plot_pca12
```

```{r pca_1_3}

#pca_with_pop$population_label <- reorder(pca_with_pop$population_label, pca_with_pop$fulani)

pca_with_pop$population_label <- factor(pca_with_pop$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca13 <- ggplot(pca_with_pop, aes(x=PC1, y=PC3, color=population_label, shape=population_label)) +
  geom_point() +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  theme(plot.title = element_text(hjust = -0.05, vjust=1),
        legend.spacing.x = unit(0.2, 'cm'),
        legend.spacing.y = unit(0.3, 'cm'),
        legend.key.size = unit(0.3,"cm"),
        legend.position = "none") +
  guides(color = guide_legend(ncol=4))
        
plot_pca13
```

```{r}

pdf("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/suppl-figure-4.pdf",
    width=10,height=7)
plot_pca12 + plot_pca13 + plot_layout(guides = 'collect') & theme(legend.position = 'bottom')
dev.off()
```

## PCA for downsampled Fulani with projection

In order to remove the strong influence of Fulani sampling with respect
to the reference populations, we carried out downsampling to \~60 Fulani
samples and have run the PCA with projection for the remaining Fulani
samples.

```{r load_data_downsampled, include=TRUE}

data_downsampled="data/pca-results/Fulani_Downsample_Projection_HD_DB-165300-190423/Fulani_Downsample_Projection_HD_DB.pca.evec"

read_table(data_downsampled, col_names = FALSE, n_max=1)[2:11] %>% 
  unlist(., use.names=FALSE) -> eigen_downsampled

eigen_downsampled <- tibble(pc = c(1:10), eigen=eigen_downsampled)

p <- ggplot(eigen_downsampled, aes(x=pc, y=eigen)) +
      geom_line(col="grey") +
      geom_point(col="dark blue") +
      theme_bw() +
      ylab("Eigen Value") +
      xlab("Principal Component")

p
```

### Load the data

```{r load_data_downsampled, include=FALSE}

pc <- as.character(paste0("PC",c(1:10)))

read_table(data_downsampled, 
           col_names = FALSE, skip=1) %>%
  dplyr::select(c(1:11)) %>%
  dplyr::rename_at(vars(as.character(paste0("X",c(1:11)))),
            ~c("id",pc)) %>% 
  tidyr::separate(id, into=c("population_label","sample_id"), sep=":") %>%
  mutate(population_label=str_remove(population_label,"_p")) -> pca_downsampled


## Read populations data
population_names <- unique(pca_downsampled$population_label)

read_tsv("data/samples/comparative_data_all_populations.tsv") %>%
  filter(population_label %in% population_names) %>%
  mutate(fulani=case_when(grepl("Fula", population_label) ~ 1,
                          grepl("Halpularen", population_label, ignore.case = TRUE) ~1,
                          region=="West Africa" ~ 2,
                          region=="Central Africa" ~ 3,
                          region=="East Africa" ~ 4,
                          region %in% c("Central Europe","North Europe","South Europe") ~ 7,
                          region=="Middle East" ~ 6,
                          region=="North Africa" ~ 5)) -> populations

pca_downsampled %>% 
  dplyr::group_by(population_label) %>%
  dplyr::summarize(n=length(sample_id),
            PC1=mean(PC1),
            PC2=mean(PC2),
            PC3=mean(PC3),
            PC4=mean(PC4),) %>%
  dplyr::inner_join(populations %>% 
               dplyr::select(population_label, longitude, latitude, lifestyle2, region, country,fulani), 
             by=c("population_label" = "population_label")) -> pca_by_pop_downsampled

pca_downsampled %>% 
  dplyr::inner_join(populations %>% 
               dplyr::select(population_label, longitude, latitude, lifestyle2, region, country,fulani), 
             by=c("population_label" = "population_label")) -> pca_with_pop_downsampled

```

### PC1 x PC2

```{r pca_1_2_downsampled}

#pca_with_pop_downsampled$population_label <- reorder(pca_with_pop_downsampled$population_label, pca_with_pop_downsampled$fulani)

pca_with_pop_downsampled$population_label <- factor(pca_with_pop_downsampled$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca12_down <- ggplot(pca_with_pop_downsampled, aes(x=PC1, y=PC2, color=population_label, shape=population_label)) +
  geom_point() +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  theme(legend.title = element_text(size=10,face="bold"),
        legend.spacing.x = unit(0.2, 'cm'),
        legend.spacing.y = unit(0.3, 'cm'),
        legend.key.size = unit(0.3,"cm")) +
  guides(color = guide_legend(ncol=2))

plot_pca12_down
```

```{r pca_1_2_downsampled_supplement}

#pca_with_pop_downsampled$population_label <- reorder(pca_with_pop_downsampled$population_label, pca_with_pop_downsampled$fulani)

pca_with_pop_downsampled$population_label <- factor(pca_with_pop_downsampled$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca12_down <- ggplot(pca_with_pop_downsampled, aes(x=PC1, y=PC2, color=population_label, shape=population_label)) +
  geom_point(size=1.5) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  ggtitle("PC1 vs PC2") +
  theme(legend.spacing.x = unit(0.2, 'cm'),
        legend.spacing.y = unit(0.3, 'cm'),
        legend.key.size = unit(0.2,"cm"),
        legend.text = element_text(size=8),
        legend.title = element_text(size=10,face="bold")) +
  guides(color = guide_legend(ncol=3))

plot_pca12_down
```

### PC1 x PC3

```{r pca_1_3_downsampled}

#pca_with_pop_downsampled$population_label <- reorder(pca_with_pop_downsampled$population_label, pca_with_pop_downsampled$fulani)

pca_with_pop_downsampled$population_label <- factor(pca_with_pop_downsampled$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca13_down <- ggplot(pca_with_pop_downsampled, aes(x=PC1, y=PC3, color=population_label, shape=population_label)) +
  geom_point(size=1) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  ggtitle("PC1 vs PC3") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = -0.05, vjust=1))

plot_pca13_down
```

### PC1 x PC4

```{r pca_1_4_downsampled}

#pca_with_pop_downsampled$population_label <- reorder(pca_with_pop_downsampled$population_label, pca_with_pop_downsampled$fulani)

pca_with_pop_downsampled$population_label <- factor(pca_with_pop_downsampled$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca14_down <- ggplot(pca_with_pop_downsampled, aes(x=PC1, y=PC4, color=population_label, shape=population_label)) +
  geom_point(size=1) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  ggtitle("PC1 vs PC4") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = -0.05, vjust=1))

plot_pca14_down
```

```{r pca_1_5_downsampled}

#pca_with_pop_downsampled$population_label <- reorder(pca_with_pop_downsampled$population_label, pca_with_pop_downsampled$fulani)

pca_with_pop_downsampled$population_label <- factor(pca_with_pop_downsampled$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca15_down <- ggplot(pca_with_pop_downsampled, aes(x=PC1, y=PC5, color=population_label, shape=population_label)) +
  geom_point(size=1) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  ggtitle("PC1 vs PC5") +
  theme(legend.position = "none",
        plot.title = element_text(hjust = -0.05, vjust=1))

plot_pca15_down
```

### PC1 x PC6

```{r pca_1_6_downsampled}

#pca_with_pop_downsampled$population_label <- reorder(pca_with_pop_downsampled$population_label, pca_with_pop_downsampled$fulani)

pca_with_pop_downsampled$population_label <- factor(pca_with_pop_downsampled$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca16_down <- ggplot(pca_with_pop_downsampled, aes(x=PC1, y=PC6, color=population_label, shape=population_label)) +
  geom_point(size=1) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  ggtitle("PC1 vs PC6") +
  theme(legend.position = "none") +
  guides(color = guide_legend(ncol=2))

plot_pca16_down
```

### PC1 x PC7

```{r pca_1_7_downsampled}

#pca_with_pop_downsampled$population_label <- reorder(pca_with_pop_downsampled$population_label, pca_with_pop_downsampled$fulani)

pca_with_pop_downsampled$population_label <- factor(pca_with_pop_downsampled$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

plot_pca17_down <- ggplot(pca_with_pop_downsampled, aes(x=PC1, y=PC7, color=population_label, shape=population_label)) +
  geom_point() +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  ggtitle("PC1 vs PC7") +
  theme_bw() +
  theme(legend.position = "none",
        plot.title = element_text(hjust = -0.05, vjust=1))

plot_pca17_down
```

```{r}

pdf("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_all_components.pdf",
    width=8,height=10)
grid.arrange(plot_pca12_down, 
             plot_pca13_down, 
             plot_pca14_down,
             plot_pca15_down,
             plot_pca16_down, ncol = 2, layout_matrix = rbind(c(1,1), c(2,3), c(4,5)))
dev.off()
```

### UMAP for down-sampled & projected dataset

##### UMAP 1 - all samples

```{r umap}

umap.config = umap.defaults

umap.config$n_neighbors=20
umap.config$min_dist=0.8
umap.config$spread=1
umap.config$n_epochs=200

pca.umap = umap(pca_with_pop_downsampled[3:12],config=umap.config, preserve.seed=TRUE)

tibble(sample_id=pca_with_pop_downsampled$sample_id, 
       population_label=pca_with_pop_downsampled$population_label, 
       region=pca_with_pop_downsampled$region,
       country=pca_with_pop_downsampled$country,
       fulani=pca_with_pop_downsampled$fulani,
       lifestyle=pca_with_pop_downsampled$lifestyle2,
       longitude=pca_with_pop_downsampled$longitude,
       latitude=pca_with_pop_downsampled$latitude,
       x=pca.umap$layout[,1], 
       y=pca.umap$layout[,2]) -> pca.umap.tibble.all
```

```{r}

pca.umap.tibble.all %>% write_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_all_20_08_1_16082023.tsv",col_names = TRUE)
```

```{r}

pca.umap.tibble.all <- read_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_all_20_08_1_16082023.tsv",col_names = TRUE)
```

```{r}

pca.umap.tibble.all$population_label <- factor(pca.umap.tibble.all$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

n_neighbors=20
min_dist=0.8
spread=1

plot_umap3 <- ggplot(pca.umap.tibble.all, aes(x=x, y=y, color=population_label,shape=population_label)) +
  geom_point(alpha=0.8,size=1) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle(paste0("UMAP: n_neighbors=",n_neighbors,"; min_dist=",min_dist,"; spread=",spread)) +
  theme(legend.position="none")

plot_umap3
```

##### UMAP 2 - all samples

```{r umap}

umap.config = umap.defaults

umap.config$n_neighbors=10
umap.config$min_dist=0.9
umap.config$spread=1
umap.config$n_epochs=200

pca.umap = umap(pca_with_pop_downsampled[3:12],config=umap.config, preserve.seed=TRUE)

tibble(sample_id=pca_with_pop_downsampled$sample_id, 
       population_label=pca_with_pop_downsampled$population_label, 
       region=pca_with_pop_downsampled$region,
       country=pca_with_pop_downsampled$country,
       fulani=pca_with_pop_downsampled$fulani,
       lifestyle=pca_with_pop_downsampled$lifestyle2,
       longitude=pca_with_pop_downsampled$longitude,
       latitude=pca_with_pop_downsampled$latitude,
       x=pca.umap$layout[,1], 
       y=pca.umap$layout[,2]) -> pca.umap.tibble.all
```

```{r}

pca.umap.tibble.all %>% write_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_all_10_09_1_16082023.tsv",col_names = TRUE)
```

```{r}

pca.umap.tibble.all <- read_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_all_10_09_1_16082023.tsv",col_names = TRUE)
```

```{r}

pca.umap.tibble.all$population_label <- factor(pca.umap.tibble.all$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

n_neighbors=10
min_dist=0.9
spread=1

plot_umap2 <- ggplot(pca.umap.tibble.all, aes(x=x, y=y, color=population_label,shape=population_label)) +
  geom_point(alpha=0.8,size=1) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle(paste0("UMAP: n_neighbors=",n_neighbors,"; min_dist=",min_dist,"; spread=",spread)) +
  theme(legend.position="none")

plot_umap2
```

##### UMAP3 - all samples

```{r umap}

umap.config = umap.defaults

umap.config$n_neighbors=10
umap.config$min_dist=0.8
umap.config$spread=1
umap.config$n_epochs=200

pca.umap = umap(pca_with_pop_downsampled[3:12],config=umap.config, preserve.seed=TRUE)

tibble(sample_id=pca_with_pop_downsampled$sample_id, 
       population_label=pca_with_pop_downsampled$population_label, 
       region=pca_with_pop_downsampled$region,
       country=pca_with_pop_downsampled$country,
       fulani=pca_with_pop_downsampled$fulani,
       lifestyle=pca_with_pop_downsampled$lifestyle2,
       longitude=pca_with_pop_downsampled$longitude,
       latitude=pca_with_pop_downsampled$latitude,
       x=pca.umap$layout[,1], 
       y=pca.umap$layout[,2]) -> pca.umap.tibble.all
```

```{r}

pca.umap.tibble.all %>% write_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_all_10_08_1_16082023.tsv",col_names = TRUE)
```

```{r}

pca.umap.tibble.all <- read_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_all_10_08_1_16082023.tsv",col_names = TRUE)
```

```{r}

pca.umap.tibble.all$population_label <- factor(pca.umap.tibble.all$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

n_neighbors=10
min_dist=0.9
spread=1

plot_umap1 <- ggplot(pca.umap.tibble.all, aes(x=x, y=y, color=population_label,shape=population_label)) +
  geom_point(alpha=0.8,size=1) +
  scale_colour_manual(values = col_palette,name="Population") +
  scale_shape_manual(values = rep(c(1:18)[-11],4)[c(1:66)],name="Population") +
  theme_bw() +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle(paste0("UMAP: n_neighbors=",n_neighbors,"; min_dist=",min_dist,"; spread=",spread)) +
  theme(legend.title=element_text(size=8,face="bold"),
        legend.text=element_text(size=8),
        legend.spacing.x = unit(0.2, 'cm'),
        legend.spacing.y = unit(0.5, 'cm'),
        legend.key.size = unit(0.3,"cm")) +
  guides(color = guide_legend(ncol=2))

plot_umap1
```

```{r}

pdf("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/umap-all-composite.pdf",width=10,height=10)
grid.arrange(plot_umap1, plot_umap2, plot_umap3, ncol = 2, 
             layout_matrix = rbind(c(1,1), c(2,3)))
dev.off()
```

#### Iterations over different UMAP parameters

```{r}

plot_umap <- function(data, n_neighbors=10, min_dist=1, spread=2){
  
  umap.config = umap.defaults
  
  umap.config$n_neighbors=n_neighbors
  umap.config$min_dist=min_dist
  umap.config$spread=spread
  
  pca.umap = umap(data[3:12],config=umap.config)
  
  tibble(sample_id=data$sample_id, 
         population_label=data$population_label, 
         longitude=data$longitude,
         latitude=data$latitude,
         region=data$region,
         country=data$country,
         x=pca.umap$layout[,1], 
         y=pca.umap$layout[,2]) -> pca.umap.tibble
  
  pca.umap.tibble %>%
    mutate(fulani=case_when(grepl("Fula", population_label) ~ "Fulani",
                            grepl("Halpularen", population_label, ignore.case = TRUE) ~"Fulani",
                            TRUE ~ region)) -> pca.umap.tibble
  
  pca.umap.tibble$fulani <- factor(pca.umap.tibble$fulani, 
                                   levels=c("Fulani","West Africa","Central Africa","East Africa","North Africa","Middle East","South Europe","Central Europe","North Europe"))
  
  p <- ggplot(pca.umap.tibble, aes(x=x, y=y, color=fulani)) +
    geom_point(alpha=0.6,size=1) +
    ggtitle(paste0("UMAP: n_neighbors=",n_neighbors,"; min_dist=",min_dist,"; spread=",spread)) +
    theme_bw() +
    theme(axis.title = element_text(size = 12,face="bold"),
          axis.text.y = element_text(size = 8),
          axis.text.x = element_text(size=8),
          legend.title=element_text(size=8,face="bold"),
          legend.text=element_text(size=8))
  
  plot(p)
  # jpeg(paste0("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/umap-",n_neighbors,"-",min_dist,"-",spread,"-Fulani_and_Reference_HD_DB.jpg"),
  #      units="in",width = 8, height = 6, res = 300, quality=100)
  # plot(p)
  # dev.off()
}

plot_umap(pca_with_pop_downsampled, n_neighbors=5, min_dist=1, spread=2)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=1, spread=2)
plot_umap(pca_with_pop_downsampled, n_neighbors=20, min_dist=1, spread=2)
plot_umap(pca_with_pop_downsampled, n_neighbors=40, min_dist=1, spread=2)

plot_umap(pca_with_pop_downsampled, n_neighbors=5, min_dist=1, spread=3)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=1, spread=3)
plot_umap(pca_with_pop_downsampled, n_neighbors=20, min_dist=1, spread=3)
plot_umap(pca_with_pop_downsampled, n_neighbors=40, min_dist=1, spread=3)

plot_umap(pca_with_pop_downsampled, n_neighbors=5, min_dist=2, spread=3)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=2, spread=3)
plot_umap(pca_with_pop_downsampled, n_neighbors=20, min_dist=2, spread=3)
plot_umap(pca_with_pop_downsampled, n_neighbors=40, min_dist=2, spread=3)

plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=0.1, spread=1)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=0.2, spread=1)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=0.5, spread=1)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=0.7, spread=1)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=0.9, spread=1)
plot_umap(pca_with_pop_downsampled, n_neighbors=10, min_dist=0.99, spread=1)
```

```{r}

plot_umap <- function(data, n_neighbors=10, min_dist=1, spread=2){
  
  umap.config = umap.defaults
  
  umap.config$n_neighbors=n_neighbors
  umap.config$min_dist=min_dist
  umap.config$spread=spread
  
  pca.umap = umap(data[3:12],config=umap.config)
  
  tibble(sample_id=data$sample_id, 
         population_label=data$population_label, 
         longitude=data$longitude,
         latitude=data$latitude,
         region=data$region,
         country=data$country,
         x=pca.umap$layout[,1], 
         y=pca.umap$layout[,2]) -> pca.umap.tibble
  
  pca.umap.tibble %>%
    mutate(fulani=case_when(grepl("Fula", population_label) ~ "Fulani",
                            grepl("Halpularen", population_label, ignore.case = TRUE) ~"Fulani",
                            TRUE ~ region)) -> pca.umap.tibble
  
  p <- ggplot(pca.umap.tibble, aes(x=x, y=y, color=fulani)) +
    geom_point(alpha=0.6,size=1) +
    ggtitle(paste0("UMAP: n_neighbors=",n_neighbors,"; min_dist=",min_dist,"; spread=",spread)) +
    theme_bw() +
    theme(axis.title = element_text(size = 12,face="bold"),
          axis.text.y = element_text(size = 8),
          axis.text.x = element_text(size=8),
          legend.title=element_text(size=8,face="bold"),
          legend.text=element_text(size=8))
  
  plot(p)
  # jpeg(paste0("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/umap-",n_neighbors,"-",min_dist,"-",spread,"-Fulani_and_Reference_HD_DB.jpg"),
  #      units="in",width = 8, height = 6, res = 300, quality=100)
  # plot(p)
  # dev.off()
}

pca_with_pop %>% filter(region=="West Africa" | region == "Central Africa" | region == "North Africa" | region=="East Africa") -> pca_with_pop_downsampled_subset

plot_umap(pca_with_pop_downsampled_subset, n_neighbors=5, min_dist=1, spread=2)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=1, spread=2)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=20, min_dist=1, spread=2)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=40, min_dist=1, spread=2)

plot_umap(pca_with_pop_downsampled_subset, n_neighbors=5, min_dist=1, spread=3)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=1, spread=3)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=20, min_dist=1, spread=3)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=40, min_dist=1, spread=3)

plot_umap(pca_with_pop_downsampled_subset, n_neighbors=5, min_dist=2, spread=3)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=2, spread=3)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=20, min_dist=2, spread=3)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=40, min_dist=2, spread=3)

plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=0.1, spread=1)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=0.2, spread=1)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=0.5, spread=1)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=0.7, spread=1)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=0.9, spread=1)
plot_umap(pca_with_pop_downsampled_subset, n_neighbors=10, min_dist=0.99, spread=1)
```

```{r}

n_neighbors=10
min_dist=0.7
spread=1

umap.config = umap.defaults
  
umap.config$n_neighbors=n_neighbors
umap.config$min_dist=min_dist
umap.config$spread=spread
umap.config$n_epochs=1000

pca.umap = umap(pca_with_pop_downsampled_subset[3:12],
                method="naive",
                config=umap.config)

tibble(sample_id=pca_with_pop_downsampled_subset$sample_id, 
       population_label=pca_with_pop_downsampled_subset$population_label, 
       longitude=pca_with_pop_downsampled_subset$longitude,
       latitude=pca_with_pop_downsampled_subset$latitude,
       region=pca_with_pop_downsampled_subset$region,
       country=pca_with_pop_downsampled_subset$country,
       lifestyle=pca_with_pop_downsampled_subset$lifestyle2,
       x=pca.umap$layout[,1], 
       y=pca.umap$layout[,2]) -> pca.umap.tibble
```

```{r}

pca.umap.tibble %>%
  mutate(fulani=case_when(grepl("Fula", population_label) ~ "Fulani",
                          grepl("Halpularen", population_label, ignore.case = TRUE) ~"Fulani",
                          TRUE ~ region)) -> pca.umap.tibble

p <- ggplot(pca.umap.tibble, aes(x=x, y=y, color=fulani,shape=lifestyle)) +
  geom_point(alpha=0.6,size=1) +
  ggtitle(paste0("UMAP: n_neighbors=",n_neighbors,"; min_dist=",min_dist,"; spread=",spread)) +
  theme_bw()

p
```

### UMAP on West and Central African Populations

```{r}

pca_with_pop %>% filter( ( region=="West Africa" | region == "Central Africa" ) & population_label != "Sudan_Nubian" ) -> pca_with_pop_downsampled_subset


pca_with_pop_downsampled_subset$population_label <- reorder(pca_with_pop_downsampled_subset$population_label, pca_with_pop_downsampled_subset$fulani)

#pca_with_pop %>% filter( ( region=="West Africa" | region == "Central Africa" ) ) -> pca_with_pop_downsampled_subset

n_neighbors=15
min_dist=0.9
spread=1
n_epochs=1000

umap.config = umap.defaults
  
umap.config$n_neighbors=n_neighbors
umap.config$min_dist=min_dist
umap.config$spread=spread
umap.config$n_epochs=n_epochs

pca.umap = umap(pca_with_pop_downsampled_subset[3:12],
                method="naive",
                config=umap.config)

tibble(sample_id=pca_with_pop_downsampled_subset$sample_id, 
       population_label=pca_with_pop_downsampled_subset$population_label, 
       longitude=pca_with_pop_downsampled_subset$longitude,
       latitude=pca_with_pop_downsampled_subset$latitude,
       region=pca_with_pop_downsampled_subset$region,
       fulani=pca_with_pop_downsampled_subset$fulani,
       country=pca_with_pop_downsampled_subset$country,
       lifestyle=pca_with_pop_downsampled_subset$lifestyle2,
       x=pca.umap$layout[,1], 
       y=pca.umap$layout[,2]) -> pca.umap.tibble
```

```{r}

## Saved result of the UMAP to be able to reproduce graphs
pca.umap.tibble %>% write_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_subset_16082023.tsv",col_names = TRUE)
```

```{r}

## Saved result of the UMAP to be able to reproduce graphs
pca.umap.tibble <- read_tsv("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/pca_umap_subset_16082023.tsv",col_names = TRUE)
```

```{r}

#pca.umap.tibble$population_label <- factor(pca.umap.tibble$population_label, levels=populations$population_label[with(populations, order(fulani, longitude))])

pca.umap.tibble %>%
  mutate(fulani=case_when(grepl("Fula", population_label) ~ "Fulani",
                          grepl("Halpularen", population_label, ignore.case = TRUE) ~"Fulani",
                          TRUE ~ region)) -> pca.umap.tibble

# cp = c(wes_palette("Cavalcanti1", 3, type = "discrete")[3],
#        wes_palette("Cavalcanti1", 3, type = "discrete")[1],
#        wes_palette("Cavalcanti1", 3, type = "discrete")[2])

plot_umap_subsistence <- ggplot(pca.umap.tibble, aes(x=x, y=y, color=fulani,shape=lifestyle)) +
  geom_point(size=1.5) +
  scale_color_manual(values=col_palette1[c(47,16,28)],name="Region") +
  scale_shape_manual(values=c(1,16),name="Subsistence") +
  xlab("UMAP1") +
  ylab("UMAP2") +
  ggtitle("") +
  theme_bw() +
  theme(legend.position = c(0.85, 0.75),
        legend.background = element_rect(fill=NA, linewidth = 0),
        legend.box.background = element_rect(fill="white",colour ="black",linewidth = 0.2),
        legend.spacing.y = unit(0,"cm"),
        legend.title=element_text(size=8,face="bold"),
        legend.text=element_text(size=8),
        legend.spacing.x = unit(0.1, 'cm'),
        legend.key.size = unit(0.4,"cm"))

plot_umap_subsistence
```

```{r}

n_neighbors=15
min_dist=0.9
spread=1

populations %>%
  filter(fulani <= 3) -> populations_subset

pca.umap.tibble$population_label <- factor(pca.umap.tibble$population_label, levels=populations_subset$population_label[with(populations_subset, order(fulani, longitude))])

plot_umap_populations <- ggplot(pca.umap.tibble, aes(x=x, y=y, color=population_label,shape=population_label)) +
  geom_point(size=1.5) +
  scale_color_manual(values=col_palette[c(1:51)],
                     name="Population") +
  scale_shape_manual(values=rep(c(1:18)[-11],3)[c(1:66)],name="Population") +
  ggtitle(paste0("UMAP: n_neighbors=",n_neighbors,"; min_dist=",min_dist,"; spread=",spread)) +
  xlab("UMAP1") +
  ylab("UMAP2") +
  theme_bw() +
  theme(legend.position="none",
        title = element_text(size=8))

plot_umap_populations
```

```{r}

pdf("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/figure2-composite.pdf",width=10,height=10)
grid.arrange(plot_pca12_down, plot_umap_populations, plot_umap_subsistence, ncol = 2, 
             layout_matrix = rbind(c(1,1), c(2,3)))
dev.off()
```

```{r}

tiff("src/rmd-reports/pca-report-Fulani_and_Reference_HD_DB/figure2-composite.tiff", units="in", width=10, height=10, res=300)
grid.arrange(plot_pca12_down, plot_umap_populations, plot_umap_subsistence, ncol = 2, 
             layout_matrix = rbind(c(1,1), c(2,3)))
dev.off()
```
